Scalar Items in Embedded Position: {A}n Experimental Revisit

### 2012-08-01 text/paper/paper1.tex
smallish changes

A number of empirical studies have already examined the availability
of readings for \as- and \es-sentences
\citep[e.g.][]{GeurtsPouscoulous2009:Embedded-Implic,CliftonDube2010:Embedded-Implic,ChemlaSpector2010:Experimental-Ev}. However,
results have not been as clear-cut as one might have hoped for: for
instance, the available empirical evidence is inconclusive as to
whether local readings exist \citep[see also][for related
discussion]{Tielvan-Tiel2012:Embedded-Scalar} and do not clearly
address all preference relations between readings relevant to
distinguish between theoretical positions. In the following, we will
argue that these heterogeneous results were partly due to the use of 
standard procedures such as picture-verification paradigm, which might not 
be sufficiently sensitive for the phenomenon under investigation,
for instance, due to possible confounding effects of the pictures used.

@@

In reaction to this situation, we therefore presented sentence
material auditorily and also employed a different kind of visual
presentation of the pictorial material.
%: subjects were presented with
%pictures that were initially covered and could incrementally be
%uncovered at the subjects' request; at each step of uncovering,
%subjects had to decide whether they could already give a truth-value
%judgement or needed more information (Comry ????)\dn{insert
%  references}. 
This way we obtained behavioral data that is both
indicative of the reading subjects assumed and independent of the
confounding pictorial effects. 
%At the same time, we hypothesized
%that the incremental nature of this task would shed light on the
%preferences over readings, because the temporal distribution of
%truth-value judgments would give away which reading subjects were
%waiting to evaluate, so to speak. 
Adding to previous studies, we made sure that our method is able 
to reveal information about reading preferences by relating our 
targets to strucutres that are known in the psycholinguistic 
literature for exhibiting certain preferences.

@@

\citet{GeurtsPouscoulous2009:Embedded-Implic} conducted a
picture-verification task to find out whether local readings of \as-
and \es-sentences are available.\fn{Actually,
  \citet{GeurtsPouscoulous2009:Embedded-Implic} did not use
  \es-sentences, but sentences where scalar \emph{some} was embedded
  under non-monotonic quantifier \emph{exactly two} (with appropriate
  pictures, of course). This case is a little more complex, but we
  will gloss over this here, treating their data, as if it was obtained
  for \es-sentences.} Subjects were presented with pictures like 
  those in Figure~\ref{fig:weak} and
\ref{fig:local} where the local reading gets a different truth-value
from the literal and the global reading. For
\as-sentences, the local reading is false for the critical picture in
Figure~\ref{fig:weak} whereas the literal and global readings are
true; for \es-sentences the local reading is true for the critical
picture in Figure~\ref{fig:local}, whereas the literal and global
readings are false.\dn{Petra wants more detail in the description of
  this study; I'm not sure that this is needed...?}

@@

\noindent Again the majority response ``both'' might speak for a
preference for the literal reading, but, as
\citeauthor{CliftonDube2010:Embedded-Implic} propose, the 17\% of
``neither'' answers in this case again suggest that the local reading
is available. Taken together,
\citeauthor{CliftonDube2010:Embedded-Implic} take these results to
contradict \citeauthor{GeurtsPouscoulous2009:Embedded-Implic}'s
findings. Local readings are, after all, attested if subjects are
given a choice as to which situation they consider most fitting for an
\as-sentence.

Taken together, the results by Clifton and Dube demonstrate that
participant' choices for \as-readings might be strongly affected by
the specific experimental paradigm used. If task demands are only
felicitous for a literal reading, the \emph{absence} of choices of
local readings cannot be accounted for by assuming an inavailability
of these readings \emph{per se}.\dn{micha does not understand this
  sentence} However, at present it is less clear whether these results
can also be generalized to further situations. For instance, subjects
rejected the \es-sentences in Geurts and Poscoulous' original study,
despite the fact that they could have accepted it based on the local
reading. It is thus necessary to test both universal and non-monotonic
quantifiers within the same experiment using a task that offers
alternatives. Additionally, it would be informative to also probe into
the availability and the relative preference of global readings. The
study of \citet{ChemlaSpector2010:Experimental-Ev} did both of that.

@@

Summing up so far, three previous studies have presented diverging
pieces of evidence concerning the status of putative local
readings. In their picture verification experiment, \citet{GeurtsPouscoulous2009:Embedded-Implic} did not observe any truth value judgments that would indicate
local readings. On the basis of this finding they argue that these readings are simply not available. On the other
hand,  \citeauthor{CliftonDube2010:Embedded-Implic} and \citeauthor{ChemlaSpector2010:Experimental-Ev} argue that the findings of  \citeauthor{GeurtsPouscoulous2009:Embedded-Implic} may be due to a bias in their experimental design and
employed different methods intended to be sensitive enough to reveal effects of local readings: either by
a choice of the situation that better befits a given sentence (\citeauthor{CliftonDube2010:Embedded-Implic}), or by a graded
truth-value judgment (\citeauthor{ChemlaSpector2010:Experimental-Ev}). Both studies found evidence in support of local readings.
To account for the diverging results we could draw a distinction between \emph{categorical} truth-value
judgments and other more sensitive measures. Firstly, it could be hypothesized
along with \citeauthor{CliftonDube2010:Embedded-Implic} and
\citeauthor{ChemlaSpector2010:Experimental-Ev} that local readings are
available, but so dispreferred that they do not affect
\emph{categorical} truth-value judgments (See, however, \citet{Crain1998} for arguments in favour of truth value judgments
as a means of detecting highly dispreferred readings).
Secondly it could also be argued, in line with traditionalism, that the apparent difference in
sensitivity of the different judgment types can be explained by the stipulation that
the results reported by \citeauthor{CliftonDube2010:Embedded-Implic}
and \citeauthor{ChemlaSpector2010:Experimental-Ev} reflect something other than
(strongly) truth-relevant speaker-intended meaning enrichments after all, so that no case
can be made for lexicalism or grammaticalism on the basis of these data \citep[c.f.][for arguments along these lines]{Tielvan-Tiel2012:Embedded-Scalar}.

@@

Be this as it may, we cannot ascribe the effects reported by  \citeauthor{CliftonDube2010:Embedded-Implic} and \citeauthor{ChemlaSpector2010:Experimental-Ev} to the availability of local readings with certainty. \citet{ChemlaSpector2010:Experimental-Ev} themeselves acknowledge
the possibility that the \emph{typicality} of a picture with respect to some meaning may affect graded truth-value
judgments. Strengthening this possibility, they show that graded judgments of
\as-sentences differ significantly for different instantiations of the
conditions in \ref{fig:AS-distinguishing-pics}, that vary in the
amount of connections without changing the assignment of truth values
to candidate readings. \citeauthor{ChemlaSpector2010:Experimental-Ev}
suggest that typicality of the pictorial material can account for
these differences, but submit that this does not explain away the
high acceptance of pictures like Figure \ref{fig:weak}. The latter point is disputed by
\citet{Tielvan-Tiel2012:Embedded-Scalar}, who demonstrates that a huge chunk of variance in the data of
\citeauthor{CliftonDube2010:Embedded-Implic} and \citeauthor{ChemlaSpector2010:Experimental-Ev} on
\as-sentences can be explained as typicality effects, dispensing the need to appeal
to the distribution of different readings. \citet{Tielvan-Tiel2012:Embedded-Scalar} elicited what he calls the typicality
structures associated with the quantifiers {\it all} and {\it some} and predicted the judgments of \as-sentences obtained by C\&S based
on these data. Thereby, he obtained an excellent model fit. We are sympathetic towards
\citeauthor{Tielvan-Tiel2012:Embedded-Scalar}'s innovative line of
reinterpretation of the data, but note, as
\citet{ChemlaSpector2010:Experimental-Ev} already observed, that his
typicality-based explanation does not extend to \es-sentences in an
obvious way\fn{For similar reasons, appeal to typicality cannot
  explain, as far as we can see, the preference for the situation in
  Figure~\ref{fig:literal-EA} over that in Figure\ref{fig:global-EA}
  given a sentence like (\ref{bsp:EA}).}.

@@

Moreover, a final omission of previous studies is that they did not
control for possible conflating effects of ``silent intonation''
(XYZ)\dn{references}. Even traditionalism maintains that local
readings are attested, but requiring a contrastive stress on the
scalar item
\citep[e.g.][]{Horn2006:The-Border-Wars,Geurts2009:Scalar-Implicat,Geurts2010:Quantity-Implic,Tielvan-Tiel2012:Embedded-Scalar}. Variation
in the interpretation of target sentences therefore could be caused by
subjects imposing different intonational patterns on the visually
presented target sentences. Ideally, we should therefore present
sentences auditorily and address directly the frequently voiced
hypothesis that local readings are triggered by contrastive
stress.\dn{relate to \citet{SchwarzClifton2008:Strengthening-o} and
  others?}

@@

\paragraph{Preference-related controls} In order to control for
artificial effects of presentation order on the availability of
readings, we included a second type of control conditions. In
particular, we tested (i), whether the order of given readings affects
preferences for ambiguous sentences and (ii), whether prosodic
information can, in principle, shift reading preferences in the
present task. Note that for our target sentences, the logical
entailment relations between readings always require local readings to
be evaluated at the end of a sequence. It could thus in principle be
possible that no subject ever reaches this point, because they gave
truth-value judgements earlier, thereby ending the trial. In that case
it would be unclear if these decisions have been affected by a general
inavailability of local readings or by the fact that one of the
earlier presented readings is the preferred one. By including globally
ambiguous structures with a known preference over available readings,
we thus intended to test whether participants in our task occasionally
choose dispreferred readings, even if they were presented subsequent 
to preferred readings.

>>>>

A number of empirical studies have already examined the availability
of readings for \as- and \es-sentences
\citep[e.g.][]{GeurtsPouscoulous2009:Embedded-Implic,CliftonDube2010:Embedded-Implic,ChemlaSpector2010:Experimental-Ev}. However,
results have not been as clear-cut as one might have hoped for: for
instance, the available empirical evidence is inconclusive as to
whether local readings exist \citep[see also][for related
discussion]{Tielvan-Tiel2012:Embedded-Scalar}, and previous studies 
did not explicitly address all preference relations between readings relevant to
distinguish between theoretical positions. In the following, we will
argue that these heterogeneous results were partly due to the use of 
standard procedures such as picture-verification paradigm, which might not 
be sufficiently sensitive for the phenomenon under investigation,
for instance, due to possible confounding effects of the pictures used.

@@

In reaction to this situation, we therefore presented sentence
materials auditorily and also employed a different kind of visual
presentation of the pictorial material.
%: subjects were presented with
%pictures that were initially covered and could incrementally be
%uncovered at the subjects' request; at each step of uncovering,
%subjects had to decide whether they could already give a truth-value
%judgement or needed more information (Comry ????)\dn{insert
%  references}. 
This way we obtained behavioral data that is both
indicative of the reading subjects assumed and independent of the
confounding pictorial effects. 
%At the same time, we hypothesized
%that the incremental nature of this task would shed light on the
%preferences over readings, because the temporal distribution of
%truth-value judgments would give away which reading subjects were
%waiting to evaluate, so to speak. 
Adding to previous studies, we made sure that our method is able 
to reveal information about reading preferences by relating our 
targets to strucutres that are known in the psycholinguistic 
literature for exhibiting certain preferences.

@@

\citet{GeurtsPouscoulous2009:Embedded-Implic} conducted a
picture-verification task to find out whether local readings of \as-
and \es-sentences are available.\fn{We will focus here on a subset 
of the conditions tested by
  \citet{GeurtsPouscoulous2009:Embedded-Implic}. Moreover, note that these authors did not use
  \es-sentences, but sentences where scalar \emph{some} was embedded
  under non-monotonic quantifier \emph{exactly two} (with appropriate
  pictures, of course). This case is a little more complex, but we
  will gloss over this here, treating their data, as if it was obtained
  for \es-sentences.} Subjects were presented with pictures like 
  those in Figure~\ref{fig:weak} and
\ref{fig:local} where the local reading gets a different truth-value
from the literal and the global reading. For
\as-sentences, the local reading is false for the critical picture in
Figure~\ref{fig:weak} whereas the literal and global readings are
true; for \es-sentences the local reading is true for the critical
picture in Figure~\ref{fig:local}, whereas the literal and global
readings are false.\dn{Petra wants more detail in the description of
  this study; I'm not sure that this is needed...?}

@@

\noindent Again, the majority response ``both'' might speak for a
preference for the literal reading, but, as
\citeauthor{CliftonDube2010:Embedded-Implic} propose, the 17\% of
``neither'' answers in this case again suggest that the local reading
is available. Taken together,
\citeauthor{CliftonDube2010:Embedded-Implic} take these results to
contradict \citeauthor{GeurtsPouscoulous2009:Embedded-Implic}'s
findings. Local readings are, after all, attested if subjects are
given a choice as to which situation they consider most fitting for an
\as-sentence.

Taken together, the results by Clifton and Dube demonstrate that
participants' choices for \as-readings might be strongly affected by
the specific experimental paradigm used. If task demands are only
felicitous for a literal reading, the \emph{absence} of choices of
local readings cannot be accounted for by assuming an inavailability
of these readings \emph{per se}.\dn{micha does not understand this
  sentence} However, at present it is less clear whether these results
can also be generalized to further situations. For instance, subjects
rejected the \es-sentences in Geurts and Poscoulous' original study,
despite the fact that they could have accepted them based on the local
reading. It is thus necessary to test both universal and non-monotonic
quantifiers within the same experiment using a task that offers
alternatives. Additionally, it would be informative to also probe into
the availability and the relative preference of global readings. The
study of \citet{ChemlaSpector2010:Experimental-Ev} did both of that.

@@

Summing up so far, three previous studies have presented diverging
pieces of evidence concerning the status of putative local
readings. In their picture verification experiment, \citet{GeurtsPouscoulous2009:Embedded-Implic} did not observe any truth value judgments that would indicate
local readings. On the basis of their results, they argue that these readings are simply not available. On the other
hand,  \citeauthor{CliftonDube2010:Embedded-Implic} and \citeauthor{ChemlaSpector2010:Experimental-Ev} argue that the findings of  \citeauthor{GeurtsPouscoulous2009:Embedded-Implic} may be due to a bias in their experimental design and
employed different methods intended to be sensitive enough to reveal effects of local readings: either by
a choice of the situation that better befits a given sentence (\citeauthor{CliftonDube2010:Embedded-Implic}), or by a graded
truth-value judgment (\citeauthor{ChemlaSpector2010:Experimental-Ev}). Both studies found evidence in support of local readings.
To account for the diverging results we could draw a distinction between \emph{categorical} truth-value
judgments and other more sensitive measures. Firstly, it could be hypothesized
along with \citeauthor{CliftonDube2010:Embedded-Implic} and
\citeauthor{ChemlaSpector2010:Experimental-Ev} that local readings are
available, but so dispreferred that they do not affect
\emph{categorical} truth-value judgments (See, however, \citet{Crain1998} for arguments in favour of truth value judgments
as a means of detecting highly dispreferred readings).
Secondly it could also be argued, in line with traditionalism, that the apparent difference in
sensitivity of the different judgment types can be explained by the stipulation that
the results reported by \citeauthor{CliftonDube2010:Embedded-Implic}
and \citeauthor{ChemlaSpector2010:Experimental-Ev} reflect something other than
(strongly) truth-relevant speaker-intended meaning enrichments after all, so that no case
can be made for lexicalism or grammaticalism on the basis of these data \citep[c.f.][for arguments along these lines]{Tielvan-Tiel2012:Embedded-Scalar}.

@@

For these reasons, we cannot ascribe the effects reported by  \citeauthor{CliftonDube2010:Embedded-Implic} and \citeauthor{ChemlaSpector2010:Experimental-Ev} to the availability of local readings with certainty. \citet{ChemlaSpector2010:Experimental-Ev} themeselves acknowledge
the possibility that the \emph{typicality} of a picture with respect to some meaning may affect graded truth-value
judgments. Strengthening this possibility, they show that graded judgments of
\as-sentences differ significantly for different instantiations of the
conditions in \ref{fig:AS-distinguishing-pics}, that vary in the
amount of connections without changing the assignment of truth values
to candidate readings. \citeauthor{ChemlaSpector2010:Experimental-Ev}
suggest that typicality of the pictorial material can account for
these differences, but submit that this does not explain away the
high acceptance of pictures like Figure \ref{fig:weak}. The latter point is disputed by
\citet{Tielvan-Tiel2012:Embedded-Scalar}, who demonstrates that a huge chunk of variance in the data of
\citeauthor{CliftonDube2010:Embedded-Implic} and \citeauthor{ChemlaSpector2010:Experimental-Ev} on
\as-sentences can be explained as typicality effects, dispensing the need to appeal
to the distribution of different readings. \citet{Tielvan-Tiel2012:Embedded-Scalar} elicited what he calls the typicality
structures associated with the quantifiers {\it all} and {\it some} and predicted the judgments of \as-sentences obtained by C\&S based
on these data. Thereby, he obtained an excellent model fit. We are sympathetic towards
\citeauthor{Tielvan-Tiel2012:Embedded-Scalar}'s innovative line of
reinterpretation of the data, but note, as
\citet{ChemlaSpector2010:Experimental-Ev} already observed, that his
typicality-based explanation does not extend to \es-sentences in an
obvious way\fn{For similar reasons, appeal to typicality cannot
  explain, as far as we can see, the preference for the situation in
  Figure~\ref{fig:literal-EA} over that in Figure\ref{fig:global-EA}
  given a sentence like (\ref{bsp:EA}).}.

@@

Finally, the above-mentioned studies did not control for possible 
conflating effects of ``silent prosody''. A growing number of experiments
suggests that numerous factors might influence accent placement and
prosodic phrasing even while reading, amongst them default accentuation,
constituent length, rhythmic phenomena, and individual variation
(Augurzky, 2008; Bader, 1998; Fodor, 2000, 2001; Kentner, 2012, Steinhauer et.
al., 2001, see also the discussion in Frazier, 2009)\dn{references}.

As it has been suggested in the theoretical literature that the availability
of a local reading might hinge on the realization of contrastive stress on the
scalar item, 
%Even traditionalism maintains that local
%readings are attested, but requiring a contrastive stress on the
%scalar item
\citep[e.g.][]{Horn2006:The-Border-Wars,Geurts2009:Scalar-Implicat,Geurts2010:Quantity-Implic,Tielvan-Tiel2012:Embedded-Scalar}, it is necessary to
control for potential variation due to potential differences in silent accent placement.
We therefore presented sentences auditorily and addressed the frequently proposed claim 
that local readings are triggered by contrastive
stress.\dn{relate to \citet{SchwarzClifton2008:Strengthening-o} and
  others?}

@@

\paragraph{Preference-related controls.} In order to control for
artificial effects of presentation order on the availability of
readings, we included a second type of control conditions. In
particular, we tested (i), whether the order of given readings affects
preferences for ambiguous sentences and (ii), whether prosodic
information can, in principle, shift reading preferences in the
present task. Note that for our target sentences, the logical
entailment relations between readings always require local readings to
be evaluated at the end of a sequence. It could thus in principle be
possible that no subject ever reaches this point, because they gave
truth-value judgements earlier, thereby ending the trial. In that case
it would be unclear if these decisions have been affected by a general
inavailability of local readings or by the fact that one of the
earlier presented readings is the preferred one. By including globally
ambiguous structures with a known preference over available readings,
we thus intended to test whether participants in our task occasionally
choose dispreferred readings, even if they were presented subsequent 
to preferred readings.
